---
title: "{{< var course.name >}}"
subtitle: |
  {{< var university.name >}}<br>
  {{< var course.semester >}}
toc: true
page-layout: full
execute:
  freeze: false
---

**This course website was last rendered {{< now "on %a %b %d, %Y at %I:%M %p" >}}.**

{{< var course.short_description >}}

## Sections and meeting locations

| Section | Instructor | Day/Time | Location | |
|----|-----|-----|-----|--|
| 1 | [Abhijit Dasgupta](mailto:abhijit.dasgupta@georgetown.edu) | Tuesdays 5:00 - 7:30 pm| ICC 116|

: {tbl-colwidths="[10,30,30,30]"}

<!-- Google Calendar Appointment Scheduling begin -->
<link href="https://calendar.google.com/calendar/scheduling-button-script.css" rel="stylesheet">
<script src="https://calendar.google.com/calendar/scheduling-button-script.js" async></script>
<script>
(function() {
  var target = document.currentScript;
  window.addEventListener('load', function() {
    calendar.schedulingButton.load({
      url: 'https://calendar.google.com/calendar/appointments/AcZssZ27KFF3HSS01h3r5RchQkEfVRRtq524UH7GEI8=?gv=true',
      color: '#039BE5',
      label: 'Book an appointment',
      target,
    });
  });
})();
</script>
<!-- end Google Calendar Appointment Scheduling -->

## Bookmark these links!

* Course website: [{{< var course.url >}}]({{< var course.url >}})
* [Syllabus, especially course policies](./site-page-content/syllabus.qmd)
* Slack Workspace: [{{< var course.slack >}}]({{< var course.slack >}})
* **Instructors only** email: [{{< var course.email >}}](mailto::{{< var course.email >}})
* {{< var university.lms >}}: [{{< var course.lms_base_url >}}{{< var course.lms_course_id >}}]({{< var course.lms_base_url >}}{{< var course.lms_course_id >}})


## Calendar

### Topics and Class Meetings

```{r}
#| echo: false
#| cache: false
source("site-page-content/class-meetings.R")

cal_df |>
  kableExtra::kbl(
    escape = F,
    col.names = c("Week", "Date", "Topic", "Labs", "Assignment", "Assignment due date", "Project milestone", "Milestone due date")
  )
# ("session_number", "monday", "session_name", "schedule_notes", "assignment_posted"

```

**Notes:**


::: callout-note
This schedule is subject to change. Check the timestamp above.
:::


* Assignments will be posted after the corresponding lecture. The due date for HW 8 is extended to account for the Thanksgiving holiday

### Deliverables


* All graded work is due at 11:59pm Eastern time on the due date
* The final project submission is due the Sunday after the last class meeting.

```{r}
#| eval: false

del_df |>
  kableExtra::kbl(
    escape = F,
    col.names = c("Due Date", "What is due")
  )
```
