---
title: "Introduction to Causal Inference"
---

## Goals for today

- Understand **why** causation matters, contrasted with correlation
- Learn the potential outcomes (counterfactual) framework
- See how confounders can mislead us—and how DAGs help
- Work through practical examples in {{< fa brands r-project >}}
- Apply propensity scores, matching, and weighting to real questions

---

## Why does causation matter? {.smaller}

We face three core challenges in observational data:

**Example 1: Does democracy increase government spending?**

- Observation: Democracies spend more on welfare
- Confounder: Wealthier countries are both more democratic and spend more
- Without causal reasoning: We misattribute spending to democracy

**Example 2: Do campaign ads change vote choice?**

- Observation: Voters who see ads support the candidate more
- Selection: Supporters self-select into watching the campaign's ads
- Without causal reasoning: We overestimate ad effectiveness

**Example 3: Does police presence reduce crime?**

- Observation: High-crime areas have more police
- Reverse causation: Police are *sent to* high-crime areas
- Without causal reasoning: We might defund police in places that need them most

---

## Correlation ≠ Causation (the usual suspects) {.smaller}

**Confounding**: A third variable X influences both treatment A and outcome Y

- Example: Education → income and health (health isn't caused by income alone)

**Reverse causation**: Y causes A, not the other way around

- Example: Does crime cause policing, or policing cause crime?

**Selection bias**: Non-random assignment to treatment

- Example: Voters who attend rallies already support the candidate

**Measurement error**: Imperfect proxies for what we care about

- Example: Survey responses vs. actual behavior

**Bias in data collection**: Systematic differences in how data is recorded

- Example: The 1948 "Dewey defeats Truman" polls (telephone sampling favored wealthy Dewey supporters)


## Types of bias

| Bias Type          | Description                                      | Example                                         |
|--------------------|--------------------------------------------------|-------------------------------------------------|
| Confounding        | A third variable affects both A and Y           | Wealth confounds democracy and spending          |
| Reverse Causation  | Y causes A instead of A causing Y                   | Crime rates influence police deployment          |
| Selection Bias     | Non-random assignment to treatment                     | Supporters self-select into watching ads         |
| Measurement Error  | Imperfect measurement of variables                     | Survey responses vs. actual behavior             |
| Data Collection Bias | Systematic differences in data recording                | Telephone polls missing non-telephone households |
---


# Counterfactuals & Potential Outcomes

## The counterfactual framework

For each unit (person, district, country), imagine two possible worlds:

- **Y(1)**: Outcome if they received treatment
- **Y(0)**: Outcome if they did not receive treatment

The **individual causal effect** for unit i:
$$\tau_i = Y_i(1) - Y_i(0)$$

(Unobservable: we only see one outcome per unit)

---

## The fundamental problem of causal inference

We observe either Y(1) or Y(0), but never both for the same unit.

| Unit | A | Y(0) | Y(1) | Observed Y | Effect |
|------|---|------|------|------------|--------|
| 1    | 1 | ?    | 10   | 10         | ? |
| 2    | 0 | 5    | ?    | 5          | ? |

**Solution**: Use aggregation and assumptions to estimate average effects instead of individual effects.

---

## Average Treatment Effects (ATE)

We want to estimate:

$$\text{ATE} = \mathbb{E}[Y(1) - Y(0)] = \mathbb{E}[Y(1)] - \mathbb{E}[Y(0)]$$

Related estimands:

- **ATT** (Average Treatment Effect on Treated): $\mathbb{E}[Y(1) - Y(0) | A=1]$  
  Effect among those who actually received treatment
  
- **CATE** (Conditional ATE): $\mathbb{E}[Y(1) - Y(0) | X=x]$  
  Effect for a specific subgroup (e.g., swing voters, rural districts)

In political science, we often care most about **ATE** for universal policies or **ATT** for targeted interventions.

---

## Three core identification assumptions

To estimate causal effects from observational data, we must assume:

**1. Consistency**  
The observed outcome equals the potential outcome under the treatment actually received.

- If unit i got A=1, then $Y_i = Y_i(1)$

**2. Positivity (overlap)**  
Every unit has a non-zero probability of receiving each treatment level.

- $0 < P(A=1 | X=x) < 1$ for all relevant x
- If some groups never get treated, we can't learn their treatment effect

**3. Conditional Exchangeability**  
After controlling for measured confounders X, treatment is as good as random.

- $(Y(1), Y(0)) \perp A | X$ (no unmeasured confounding)
- This is the hardest assumption—DAGs help us think through it

---

## Directed Acyclic Graphs (DAGs)

A DAG shows causal relationships between variables. We use them to reason about confounding and identify adjustment sets.

```{r}
#| label: dag-intro
#| layout-ncol: 2
library(dagitty)
library(ggdag)
library(tidyverse)

# Simple confounding: X confounds the A → Y relationship
g_conf <- dagitty('dag { X -> A; X -> Y; A -> Y }')
ggdag(g_conf) + 
  ggtitle('Confounding:\nX → A and X → Y') + 
  theme_void() + 
  theme(plot.title = element_text(size=14, hjust=0.5))

# Mediator: M is the mechanism through which A affects Y
g_med <- dagitty('dag { A -> M -> Y; A -> Y }')
ggdag(g_med) + 
  ggtitle('Mediation:\nA → M → Y') + 
  theme_void() + 
  theme(plot.title = element_text(size=14, hjust=0.5))
```

**Rule**: Adjustment sets block backdoor paths while leaving front-door paths open.

---

## Backdoor criterion

A **backdoor path** from A to Y starts with an arrow pointing *into* A:

$$A \leftarrow \cdots \rightarrow Y$$

To estimate the causal effect of A on Y, we must **block all backdoor paths**.

```{r}
#| label: backdoor-example
g_backdoor <- dagitty('dag { A <- X -> Y; A -> Y }')
ggdag(g_backdoor) + 
  ggtitle('Backdoor path: A ← X → Y\n(confounding)') + 
  theme_void() + 
  theme(plot.title = element_text(size=14, hjust=0.5))

# Find minimal adjustment set
adjustmentSets(g_backdoor, exposure = 'A', outcome = 'Y')
```

By conditioning on X, we block the backdoor path and isolate the causal effect.

---

## What is a propensity score?

The **propensity score** is the probability of receiving treatment given observed covariates:

$$e(X) = P(A=1 | X)$$

**Key insight**: Units with the same propensity score are balanced on X in expectation.  
Comparing treated and control units with similar scores gives an unbiased estimate of the treatment effect.

**Three ways to use propensity scores**:

1. **Matching**: Pair treated and control units with similar scores
2. **Weighting (IPTW)**: Weight each unit inversely by score to create a pseudo-population where treatment is independent of X
3. **Stratification**: Compare treated and control within score "bins"

**Political example**: Do swing voters (same propensity to support either candidate) change their vote based on negative ads?

---

## Simulation example: Setup

We'll simulate a dataset where treatment assignment depends on covariates and the treatment effect varies across units (heterogeneity).

```{r}
#| label: sim-data
set.seed(2025)
N <- 2000
X1 <- rnorm(N)           # Continuous covariate (e.g., ideology)
X2 <- rbinom(N, 1, 0.4)  # Binary covariate (e.g., urban vs. rural)

# True heterogeneous treatment effect: τ(x) = 1 + 0.5 * X1
tau <- 1 + 0.5 * X1

# Propensity score depends on X1 and X2
ps <- plogis(-0.2 + 0.8 * X1 - 0.5 * X2)
A <- rbinom(N, 1, ps)

# Potential outcomes
Y0 <- 2 + 0.5 * X1 - 0.3 * X2 + rnorm(N, 0, 1)
Y1 <- Y0 + tau

# Observed outcome
Y <- ifelse(A == 1, Y1, Y0)

df <- tibble(Y, A, X1, X2, ps, tau)
head(df, 10)
```

---

## Naive estimate (biased)

Simply comparing treated vs. control ignores confounding:

```{r}
#| label: naive-estimate
naive_ate <- df |>
  group_by(A) |>
  summarize(mean_Y = mean(Y), .groups = 'drop') |>
  summarize(naive_ATE = diff(mean_Y))

naive_ate
```

This is biased because treated units have higher X1 on average, which independently increases Y.

---

## Regression adjustment

Control for confounders in an outcome model:

```{r}
#| label: regression-adjustment
reg_mod <- lm(Y ~ A + X1 + X2, data = df)

broom::tidy(reg_mod) |>
  filter(term == 'A') |>
  select(term, estimate, std.error, p.value)
```

This isolates the effect of A after adjusting for X1 and X2.

---

## Inverse Probability of Treatment Weighting (IPTW)

Estimate propensity scores, then weight each unit by the inverse of their probability of receiving their observed treatment:

```{r}
#| label: iptw
# Estimate propensity score
ps_mod <- glm(A ~ X1 + X2, family = binomial(), data = df)

df <- df |>
  mutate(
    ps_hat = predict(ps_mod, type = 'response'),
    # Weight: 1/P(A=1) if treated, 1/P(A=0) if control
    w = ifelse(A == 1, 1 / ps_hat, 1 / (1 - ps_hat))
  )

# Weighted outcome regression
iptw_mod <- lm(Y ~ A, weights = w, data = df)

broom::tidy(iptw_mod) |>
  filter(term == 'A') |>
  select(term, estimate, std.error, p.value)
```

This creates a pseudo-population where treatment is independent of X.

---

## Propensity Score Matching

Match treated units to control units with similar propensity scores:

```{r}
#| label: matching
library(MatchIt)

# 1:1 nearest-neighbor matching on propensity score
m_out <- matchit(A ~ X1 + X2, data = df, method = 'nearest', ratio = 1)

# Extract matched sample
matched <- match.data(m_out)

# Estimate ATE on matched sample
matched_ate <- matched |>
  group_by(A) |>
  summarize(mean_Y = mean(Y), .groups = 'drop') |>
  summarize(matched_ATE = diff(mean_Y))

matched_ate
```

Matching directly creates a balanced sample where treated and control are comparable.

---

## Heterogeneous Treatment Effects (HTE)

Treatment effects may vary across subgroups. A simple approach: include interactions.

```{r}
#| label: hte-interaction
# Outcome model with interaction
int_mod <- lm(Y ~ A * X1 + X2, data = df)

# CATE at different X1 values
cate_fn <- function(x_val) {
  coefs <- coef(int_mod)
  # CATE(x) = β_A + β_(A:X1) * x
  ate_x <- coefs['A'] + coefs['A:X1'] * x_val
  tibble(X1_value = x_val, CATE = ate_x)
}

bind_rows(
  cate_fn(-1),  # Low ideology
  cate_fn(0),   # Moderate
  cate_fn(1)    # High ideology
)
```

Treatment effects are stronger for high-X1 units (effect of 1.5 vs. 0.5).

---

## Summary: When to use each method

| Method | When to use | Pros | Cons |
|--------|------------|------|------|
| **Regression** | Few confounders, linear effects | Simple, interpretable | Assumes linearity; vulnerable to model misspecification |
| **IPTW** | Moderate confounding | Flexible; creates interpretable pseudo-population | Weights can be extreme; requires positivity |
| **Matching** | Need transparent comparisons | Creates balanced sample; easy to diagnose | May discard data; harder with many confounders |
| **HTE methods** | Interested in subgroups | Answers heterogeneous effects | Requires more data; multiple testing concerns |

---

## Causal inference in political science: Key applications

1. **Policy evaluation**: Did a law reduce unemployment, or was it passed during recovery?

2. **Campaign effects**: Do campaign ads change votes, or do they target supporters?

3. **Representation**: Does descriptive representation (diversity) cause better policy outcomes?

4. **International relations**: Does trade reduce conflict, or do peaceful countries trade more?

All require causal thinking + DAG reasoning + careful estimation.

---

## Key takeaways

✓ Correlation ≠ causation. Confounding, selection bias, and reverse causation are everywhere.

✓ **Counterfactuals** let us formalize what we mean by "causal effect."

✓ **DAGs** help us reason about confounding and decide what to adjust for.

✓ **Propensity scores** create comparable groups and estimate treatment effects.

✓ **Heterogeneous effects** matter: treatment doesn't affect everyone the same way.

✓ No method is perfect. Always report robustness checks and acknowledge assumptions.

---

## References & Further Reading

- Hernán, M. A., & Robins, J. M. (2020). *Causal Inference: What If*. Chapman and Hall/CRC.  
  [Free online at: www.hsph.harvard.edu/miguel-hernan/causal-inference-book/](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)

- Imbens, G. W., & Rubin, D. B. (2015). *Causal Inference in Statistics, Social, and Biomedical Sciences*. Cambridge University Press.

- Pearl, J. (2009). *Causality: Models, Reasoning and Inference*. Cambridge University Press.

- Cunningham, S. (2021). *Causal Inference: The Mixtape*. [Free online at: mixtape.scunning.com](https://mixtape.scunning.com)

- Wickham, H., & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media.
