---
title: "Categorical variables and factors in R"
filters:
  - quarto
  - line-highlight
---

```{r}
#| include: false
library(tidyverse)
```

## What are categorical variables?

Categorical variables are variables that

+ have values defining categories of things
+ typically have a few unique values
+ may or may not be ordered
+ are not interval-scaled, i.e., their differences don't make sense _per se_


::: {.columns}
::: {.column}
## Non-ordered

1. Race (White, Black, Hispanic, Asian, Native American)
1. Gender (Male, Female, Other)
1. Geographic regions (Africa, Asia, Europe, North America, South America)
1. Genes/Proteins
:::
::: {.column}
## Ordered

1. Income levels (< $10K, $10K - $25K, $25K - $75K, $75K - $100K)
1. BMI categories (Underweight, Normal, Overweight, Obese)
1. Number of bedrooms in houses (1 BR, 2BR, 3BR, 4BR)
:::
::: 


## The `factor` data type

R stores categorical variables as type `factor`.

+ You can coerce a character or numeric object into a factor using `as.factor`.
+ You can check if an object is a factor with `is.factor`.
+ You can create a factor with the function `factor`.


::: {.callout-note}
## The `factor` function
 `factor(x = character(), levels, labels = levels,
      exclude = NA, ordered = is.ordered(x), nmax = NA)`

 factor returns an object of class "factor" which has a set of integer codes the length of x with a "levels" attribute of mode character and unique
:::

+ Internally, each level of a factor is coded as an integer
+ Each such integer has a corresponding `level` which is a character, describing the level. 
+ You can add `labels` to each level to change the printed form of the factor. 


```{r factors-1}
#| code-fold: show 
x <- c('Maryland','Virginia', 'District', 'Maryland','Virginia')  # a character vector
xf <- as.factor(x)
xf
```

There are three levels, that by default are in alphabetical order

::: {.columns}
::: {.column}
```{r factors-2}
#| code-fold: show 
as.integer(xf)
```

+ District = 1, Maryland = 2, Virginia = 3
:::
::: {.column}
```{r factors-3}
#| code-fold: show 
as.character(xf)
```

+ Get original characters back
:::
:::


Now let's look at what happens when we convert a numeric vector to a factor

```{r factors-4}
#| code-fold: show 
y <- c(5, 3, 9, 4, 5, 3)
yf <- as.factor(y)
yf
```

Levels are still in alphanumeric order

::: {.columns}
::: {.column}
```{r factors-5}
#| code-fold: show 
as.numeric(yf)
```

+ Note, we don't get original integers back!!
+ 3 = 1, 4 = 2, 5 = 3, 9 = 4
:::
::: {.column}
```{r factors-6}
#| code-fold: show 
as.numeric(as.character(yf))
```

+ This is how you get numbers back
:::
::: 




::: {.columns}
::: {.column}
```{r factors-7}
#| code-fold: show 
x <- c('MD','DC','VA','MD','DC')
xf <- factor(x)
unclass(xf)
```

:::
::: {.column}
```{r factors-8}
#| code-fold: show 
x <- c('MD','DC','VA','MD','DC')
xf <- factor(x, levels = c('MD','DC','VA'))
unclass(xf)
```
:::
::: 

+ If I change the level designation, the underlying coding changes
+ This is important when a factor is an independent variable in a regression model



The `drv` variable in the `mpg` dataset tells us the kind of drive (front, rear or 4-wheel) each car has. However
it's coded as `f`, `r`, and `4`, which is not great for display purposes. We can re-label these levels, but we have to be a bit careful

:::: .columns
::: .column
```{r factors-9, error=TRUE}
#| code-fold: show 
x <- mpg$drv
xf <- factor(x, 
              levels = c('4-wheel','Front wheel', # <1>
                         'Rear wheel'))
head(xf)
```
1. levels have to match what's actually in the data
:::
::: .column
```{r factors-10}
#| code-fold: show 
#| source-line-numbers: "3-5"

x <- mpg$drv
xf <- factor(x, 
              levels = c('4', 'f','r'), # <1>
              labels = c('4-wheel','Front wheel', #<2>
                         'Rear wheel'))
head(xf)
```
1. levels have to match what's actually in the data
2. labels are what you want the levels to be displayed as
:::
::::


Levels have to match what's actually in the original data, but you can re-label the levels.


# Why factors?
## Factors are R's discrete data type

+ Factors are interpreted as discrete by R's functions

::: {.columns}
::: {.column}
```{r factors-11, message=TRUE, warning=TRUE, fig.height=3}
ggplot(mpg,
       aes(year, cty))+
  geom_boxplot()
```
:::
::: {.column}
```{r factors-12, fig.height=3}
ggplot(mpg, 
       aes(as.factor(year), cty))+
  geom_boxplot()
```
:::
::: 

## Dummy variables: converting factors to numeric for modeling

Modeling is a numerical enterprise, where we optimize a function of numerical variables to fit a numerical outcome. 
So categorical variables need to be converted to some numerical representation. The most common approach is to use
_dummy variables_ (also called indicator variables or one-hot encoding). With this approach, if a factor has _n_ levels, we create _n-1_ new binary variables, each taking on values 0 or 1. Each new variable indicates whether the observation belongs to that level or not. The level corresponding to integer code 1 is omitted as the reference level.

For example, if we have a factor `race` with levels White, Black, Hispanic, Asian, Native American, we can create 4 new binary variables `black`, `hispanic`, `asian`, `native_american`, where each variable takes on value 1 if the observation
belongs to that race, and 0 otherwise. White is the reference level, so if all 4 new variables are 0, then the observation is White.

::: {.callout-caution}
## Why don't we just code categories as integers, e.g., 1, 2, 3?
Coding categories as integers implies an ordering and distance between the categories that may not exist. For example, if we code race categories as 1 = White, 2 = Black, 3 = Hispanic, 4 = Asian, 5 = Native American, then we are implying that the difference between White and Black is the same as the difference between Hispanic and Asian. This is not true, and can lead to misleading results in models. There is no reasonable interpretation of this difference. 
:::
## Dummy variables are automatically created from factors

:::: .columns
::: .column
```{r factors-13, eval=F}
#| code-fold: show 
model.matrix(~species, data = palmerpenguins::penguins)
```
```{r factors-14, echo = F}
palmerpenguins::penguins %>% 
  group_by(species) %>% 
  slice_head(n=2) %>% 
  ungroup() %>% 
  model.matrix(~species, data = .)
```
:::
::: .column
+ If a factor has _n_ levels, you get _n-1_ dummy variables
+ The level corresponding to integer code 1 is omitted as the reference level
:::
::::

Changing the base level (integer code 1) changes model interpretation since
it changes the reference level against which all other levels are compared.

This enconding is automatically done when a factor is used as an independent variable in a model.

::: {.aside}
**Note on contrasts**  
R uses _treatment contrasts_ by default, which is the dummy variable approach described above. Other contrast schemes are possible, such as sum contrasts, Helmert contrasts, polynomial contrasts, etc. See `?contrasts` for more details.
:::



# Manipulating factors<br/><br/>The `forcats` package (part of `tidyverse`)



## Effect in models

::: {.columns}
::: {.column}
```{r factors-15}
library(palmerpenguins)
m <- lm(body_mass_g ~ species, data = penguins)
broom::tidy(m)
```
Compare with Adele
:::
::: {.column}
```{r factors-16}
p1 <- penguins %>% 
  mutate(species = fct_relevel(species, 'Gentoo'))
m1 <- lm(body_mass_g ~ species, data=p1 )
broom::tidy(m1)
```
Compare with Gentoo
:::
:::

Providing only one level to `fct_relevel` makes that the base level (integer code 1).  
You can also fully specify all the levels in order, or partially specify them. If you 
partially specify them, the remaining levels will be put in alphabetical order after the
ones you specify.

## Effect in plots

::: {.columns}
::: {.column}
```{r factors-17, fig.height=4}
ggplot(penguins, 
       aes(x = species))+
  geom_bar()
```

:::
::: {.column}
```r
ggplot(p1, 
       aes(x = species))+
  geom_bar()
```

Changes the order in which bars are plotted
:::
:::

## Extra levels

::: {.columns}
::: {.column}
```{r factors-19}
x <- factor(str_split('statistics', '')[[1]], 
            levels = letters)
x
```
```{r factors-20}
p1 <- penguins %>% filter(species != 'Gentoo')
fct_count(p1$species)
```

:::
::: {.column}
```{r factors-21}
fct_drop(x)
```
```{r factors-22}
p1 <- p1 %>% mutate(species = fct_drop(species))
fct_count(p1$species)
```
:::
::: 

Getting rid of extra levels<br/><br/>
Sometimes levels with no data show up in summaries or plots

# Ordering levels by frequency

::: {.columns}
::: {.column}
```{r factors-23, fig.height=4} 
ggplot(penguins, 
       aes(x = species))+
  geom_bar()
```

:::
::: {.column}
```{r factors-24, fig.height=4}
ggplot(penguins,
       aes(x = fct_infreq(species)))+
  geom_bar()
```

:::
::: 

Ordering levels from most to least frequent



## Ordering levels by values of another variable

::: {.columns}
::: {.column}
```{r factors-25, fig.height=4}
ggplot(penguins, 
       aes(x = species,
           y = bill_length_mm))+
  geom_boxplot()
```
:::
::: {.column}
```{r factors-26, fig.height=4}
ggplot(penguins, 
       aes(x = fct_reorder(species, bill_length_mm, 
                           .fun=median, na.rm=T),
           y = bill_length_mm))+
  geom_boxplot() + labs(x = 'species')
```

:::
:::

`fct_reorder` is useful for ordering a plot by ascending or descending levels. This 
makes the plot easier to read. 


## Ordering levels by values of another variable

```{r factors-27, echo=FALSE}
if('USArrests' %in% ls()) rm(USArrests)
if(!('State' %in% names(USArrests))) USArrests <- USArrests %>% rownames_to_column('State')

```


```{r factors-28, eval=F}
USArrests <- USArrests %>% rownames_to_column('State')
```


::: {.columns}
::: {.column}
```{r factors-29, fig.height=5}
ggplot(USArrests, aes(x=State, y = Murder))+
  geom_bar(stat = 'identity') + 
  theme(axis.text = element_text(size=6))+
  coord_flip()
```

:::
::: {.column}
```{r factors-30, fig.height=5}
ggplot(USArrests, aes(
  x = fct_reorder(State, Murder),
  y = Murder))+
  geom_bar(stat = 'identity')+
  theme(axis.text = element_text(size=6))+
  coord_flip()
```
:::
:::


## Order levels based on last values when plotting 2 variables

The level ordering also shows up in the order of levels in the legends of plots. 
Suppose you are plotting two variables, grouped by a factor.

::: {.columns}
::: {.column}
```{r factors-31, fig.height=4}
ggplot(iris, aes(
  x = Sepal.Length, 
  y = Sepal.Width, 
  color = Species))+
  geom_smooth(se=F)
```

:::
::: {.column}
```{r factors-32, fig.height=4}
ggplot(iris, aes(
  x = Sepal.Length, 
  y = Sepal.Width, 
  color = fct_reorder2(Species,
                       Sepal.Length, Sepal.Width)))+
  geom_smooth(se=F) + labs(color = 'Species')
```

:::
:::



## Further exploration

1. [forcats cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/factors.pdf)
1. [Chapter 15](https://r4ds.had.co.nz/factors.html) of R4DS
