# Base maps

## What are base maps?

Base maps are (often raster) representations of geographic entities (what we usually call "maps") upon which we can then
layer other data based on location.

- Think Waze or Google Maps or Apple Maps

The underlying geographic information is used for reference and context, and helps with interpreting the
data being presented

We'll often prefer interactive maps here (what we're used to from GPS), and these are most often
done using the `leaflet.js` package and it's various wrappers (`leaflet` in R and `folium` in Python)

## Plotting Houston crime

```{r}
#| echo: true
#| code-fold: true
#| cache: true
#| fig-width: 3
#| fig-height: 3
#| label: houston
library(osmdata)
library(sf)
#houston = getbb("Houston") |> opq() |>
#  add_osm_feature(key ='highway', value = c('motorway','primary','secondary')) |>
#  osmdata_sf()
#houston_crime = st_as_sf(ggmap::crime |> drop_na(lon, lat),
#  coords = c('lon','lat'),
#  crs = st_crs(houston$osm_lines))
#save(houston_crime, houston, file='topics/geospatial/data/houston.rda')
load('topics/geospatial/data/houston.rda')
ggplot() +
  geom_sf(data = houston$osm_polygons) +
  geom_sf(data = houston$osm_lines) +
  geom_sf(data = filter(houston_crime, offense == "murder"), color='red', size=2)+
  coord_sf(xlim = getbb("Houston")[1,], ylim = getbb("Houston")[2,]) +
  labs(title = "Murders in Houston, 2010")

```
## Plotting Houston crime

```{r}
#| label: leaflet1
library(leaflet)
crime_murder = ggmap::crime |> drop_na(lon, lat) |> filter(offense == "murder")
leaflet() |>
  addTiles() |>
  addMarkers(lng = crime_murder$lon, lat = crime_murder$lat)
```

## Visualizing NYC subway stations

```{python}
# | label: geopandas1
# | code-fold: true
# | message: false
import geopandas
import folium

gdf = geopandas.read_file(
    "https://raw.githubusercontent.com/python-visualization/folium-example-data/main/subway_stations.geojson"
)
gdf["href"] = '<a href="' + gdf.url + '">' + gdf.url + "</a>"
gdf["service_level"] = gdf.notes.str.split(", ").apply(
    lambda x: len([v for v in x if "all" in v])
)
gdf["lines_served"] = gdf.line.str.split("-").apply(lambda x: len(x))
service_levels = gdf.service_level.unique().tolist()
colors = ["orange", "yellow", "green", "blue"]
m = folium.Map(location=[40.75, -73.95], zoom_start=13)

folium.GeoJson(
    gdf,
    name="Subway Stations",
    marker=folium.Circle(
        radius=4, fill_color="orange", fill_opacity=0.4, color="black", weight=1
    ),
    tooltip=folium.GeoJsonTooltip(fields=["name", "line", "notes"]),
    popup=folium.GeoJsonPopup(fields=["name", "line", "href", "notes"]),
    style_function=lambda x: {
        "fillColor": colors[x["properties"]["service_level"]],
        "radius": (x["properties"]["lines_served"]) * 30,
    },
    highlight_function=lambda x: {"fillOpacity": 0.8},
    zoom_on_click=True,
).add_to(m)

m


```
