---
title: "Brackets in R"
execute:
  echo: true
  warning: false
  error: false
  message: false
  eval: false
---

There are three kinds of brackets in R


### `[]` are used for extracting elements from arrays, matrices, data frames.

::: {.panel-tabset}

## Arrays

- `x[3]` is the 3rd element of an array `x`
- `d[1,3]` is the element in the 1st row and 3rd column of a matrix/data frame `d`
- `d[2,]` is the entire first __row__ of a matrix/data frame `d`
- `d[,4]` is the entire 4th __column__ of a matrix/data frame `d`
- `d[,'gender']` is the column named _gender_ in a _data frame_ `d`

## Lists

To extract single elements, you use a .acidinline[double bracket:]
- `lst[[3]]` is the third element of the list `lst`
- `lst[['apple']]` is the element named _apple_ in a named list `lst`

To extract a sub-list, you use .acidinline[single brackets]:
+ `lst[1:3]` creates a sub-list with the first 3 elements of `lst`
:::


### `()` are used for specifying arguments to functions

- `mean(x)` gives the mean of an array of numbers `x`
- `summary(d)` gives a summary representation of a data frame `d`
- There are several functions used to make the following visualization

```{r, eval=F}
ggplot(beaches, aes(x = temperature, y = rainfall)) + 
  geom_point() + 
  geom_smooth() + 
  theme_classic()
```


### `{}` are used to contain groups of commands/statements

:::: {.columns}
::: {.column}
A conditional statement

```{r Rintro-1, eval = F}
if (age < 18){
   person <- 'Minor'
} else if (age > 65) {
  person <- 'Senior'
} else {
  person <- 'Adult'
}
```
:::
::: {.column}
A function definition

```{r}

my_mean <- function(x, na.rm = T){
  if(na.rm){
    x <- x[!is.na(x)]
  }
  s <- sum(x)
  n <- length(x)
  mn <- s / n # There is a built-in function mean, so I don't use that word
  return(mn)
}
```
:::
::::
