## Syntactic names

Following these rules will always give you valid object names without the need for additional trouble.

1. Contains letters, numbers, `.` or `_` characters
1. Starts with a letter or `.` not followed by a number
1. Is not a reserved word ()

. . .

So `.2way` is not a syntactically valid name


## Going wild

You can actually use **any** name for an object, as long as you encase it in backticks

The name can then include spaces, special characters, start with a number, etc.

Still, don't use reserved words.

--

:::: {.columns}
::: {.column}
You can do 

`` `First variable` ``

`` `2 many bikes` ``
:::
::: {.column}
[But should you?!!]{style="color:red; font-size:150%"}
:::
:::: 



## Naming conventions

There are several prevalent naming conventions in R

+ snake case, e.g., `my_variable`
+ lower camel case, e.g., `myVariable`
+ upper camel case, e.g., `MyVariable`
+ period separated, e.g., `my.variable`
+ lower case, e.g., `myvariable`

Choose one and stick with it. It will save you some sanity.


> The **janitor** package has a nice function called `clean_names` which will 
transform the column names in a data.frame to follow a particular naming convention, 
with the default being snake case.

## Resources

The following commands, typed in the R console, will get you to the corresponding help files.
+ `?make.names`
+ `?assign`


# Brackets


## Brackets

There are three kinds of brackets in R


`[]` are used for extracting elements from arrays, matrices, data frames.


::: {.panel-tabset}

## Arrays

- `x[3]` is the 3rd element of an array `x`
- `d[1,3]` is the element in the 1st row and 3rd column of a matrix/data frame `d`
- `d[2,]` is the entire first __row__ of a matrix/data frame `d`
- `d[,4]` is the entire 4th __column__ of a matrix/data frame `d`
- `d[,'gender']` is the column named _gender_ in a _data frame_ `d`

## Lists

To extract single elements, you use a .acidinline[double bracket:]
- `lst[[3]]` is the third element of the list `lst`
- `lst[['apple']]` is the element named _apple_ in a named list `lst`

To extract a sub-list, you use .acidinline[single brackets]:
+ `lst[1:3]` creates a sub-list with the first 3 elements of `lst`
:::

## Brackets


`()` are used for specifying arguments to functions

- `mean(x)` gives the mean of an array of numbers `x`
- `summary(d)` gives a summary representation of a data frame `d`
- There are several functions used to make the following visualization

```{r, eval=F}
ggplot(beaches, aes(x = temperature, y = rainfall)) + 
  geom_point() + 
  geom_smooth() + 
  theme_classic()
```

## Brackets


`{}` are used to contain groups of commands/statements

:::: {.columns}
::: {.column}
A conditional statement

```{r Rintro-1, eval = F}
if (age < 18){
   person <- 'Minor'
} else if (age > 65) {
  person <- 'Senior'
} else {
  person <- 'Adult'
}
```
:::
::: {.column}
A function definition

```{r Rintro-2, eval=F}
my_mean <- function(x, na.rm = T){
  if(na.rm){
    x <- x[!is.na(x)]
  }
  s <- sum(x)
  n <- length(x)
  mn <- s / n # There is a built-in function mean, so I don't use that word
  return(mn)
}
```
:::
:::: 


